---
title: "Dealing with Data and Packages"
date: "May 7, 2019"
output: 
  html_document:
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: true
    df_print: paged
    theme: spacelab
  revealjs::revealjs_presentation:
    fig_width: 14
    fig_height: 7
    self_contained: true
    theme: black
    highlight: zenburn
    css: slidesdjh.css
    center: false
    transition: slide
    reveal_options:
      controls: true
      progress: false
      width: 1080
      height: 600
---

# Extending R with packages {data-background="https://pbs.twimg.com/media/CvzEQcfWIAAIs-N.jpg" data-background-size="contain"}

## Packages in R

>- So far, we have been using "base" R.
>- One of the most compelling reasons to use R is the enormous ecosystem of packages that extend its capabilities.
>- [CRAN task view](https://cran.rstudio.com/web/views/)

## Install the Tidyverse

We will be using a suite of packages designed for data science.

**$\Rightarrow$** Install the `tidyverse` by executing the following command in the console:

```{r, eval=FALSE}
install.packages("tidyverse")
```

This installation may take a few moments.

# *Getting Data into R* {data-background="https://tr2.cbsistatic.com/hub/i/2016/05/27/b6dd6d05-84b0-4eeb-ad7f-ddea46d0f15c/unexpected-error.jpg" data-background-size="contain"}

## Reproducibilty principles

>- We want other people to be able to look at our data and reproduce our statistical analyses.
>- We should make our data public.
>- We should make our code public.
>- Importing data into a statistical package often involves choices. These choices should be reproducible.
>- Therefore, **our code should import our data.** Avoid using menus or mouse clicks to import data.

## Importing Data from a website

- In the [last lesson](intro_stats.html), we read most of our data sets off of websites.
- For many R commands, `file` can be a path to a local file, or a URL.
- This is fairly reproducible. (It would be better if the code and data were in the same place.)

## **$\Rightarrow$** Make a new project

1. In the *File* menu, choose *New Project...*. In the dialog that appears, select *New Directory* followed by  *New Project*.

2. For the *Directory Name*, use `DataWorkshop`. Click on *Browse* and find a subdirectory where you want this new directory to reside (e.g., Documents, Dropbox, etc.). You will need to be able to find this directory later.

3. Click on *Create Project*. A new empty project should load. A file named `DataWorkshop.Rproj` should appear in the *Files* tab in the lower right pane.

## **$\Rightarrow$** Make a new RMarkdown document

1. In the *File* menu, choose *New File > RMarkdown...*.
2. In the "New R Markdown" dialog box, give the document a title. Keep the other default settings (Document, HTML).
3. Clicking OK will open a new RMarkdown document with some boilerplate code. *Delete* everything below line 7. 
4. *Save* the document as `datatidy.Rmd` in the project directory `DataWorkshop`. After you save the file, it should appear in the *Files* tab, along with the file `DataWorkshop.Rproj`.

## Importing Data from a local file

- Quite often, we have data in a local file we want to use.
- We should put our data where our code is: the project folder.

## **$\Rightarrow$** Download some data

1. Download the Excel file [flights.xlsx](http://math.westmont.edu/data/flights.xlsx) to your computer. **Make sure it ends up in the project folder `DataWorkshop` that you just created.** Depending on your browser settings, it may end up in your `Downloads` folder, and you will have to move it to the `DataWorkshop` folder. 
2. Refresh the view in the *Files* pane of your project, and make sure that the `flights.xlsx` file is there.

## **$\Rightarrow$** Read the Excel file and view 3 ways

1. Execute the following command in the terminal. The first line loads the tidyverse package `readxl`, and the second line reads the entire spreadsheet into the `flights` variable.

```{r}
library(readxl)
flights <- read_excel("flights.xlsx")
```

2. Type `flights` in the terminal.

3. In the *Environment* tab in the upper-right corner, click on the triangle next to the word `flights`.
4. In the *Environment* tab in the upper-right corner, click on the word `flights`.

## Tibbles

- The variable `flights` is a `tibble`, which is a data frame with additional features.
- Tibbles are part of the tidyverse.
- Tibbles can be treated just like data frames.

# *Shaping Data* {data-background="https://laughingsquid.com/wp-content/uploads/2017/10/game-of-thrones-timelapse-of-an-artist-sculpting-and-airbrushing-a-bust-of-the-night-king.gif" data-background-size="contain"}

## Shaping your data with dplyr

The remainder of this lesson is adapted from the [Data transformation](https://r4ds.had.co.nz/transform.html) chapter in *R for Data Science*.

## Dplyr basics: Five functions

- Pick observations by their values (`filter`).
- Reorder the rows (`arrange`).
- Pick variables by their names (`select`).
- Create new variables with functions of existing variables (`mutate`).
- Collapse many values down to a single summary (`summarise`).

## Load Dplyr

**$\Rightarrow$** Load the `dplyr` package in the console.

```{r}
library(dplyr)
```


## Filter

Select observations based on their values.

```{r}
flights %>% filter(arr_delay > 100)
```

## Filter: Multiple conditions (AND)

```{r}
flights %>% filter(arr_delay > 100, day == 1)
```

## Filter: Multiple conditions (OR)

```{r}
flights %>% filter(arr_delay > 100 | day == 2)
```

## **$\Rightarrow$** Filter exercises

1. Find all flights that flew to or from Houston (IAH or HOU).

2. Find all flights that had a departure delay of more than one hour, but also made up at least 15 minutes during the flight.

## Arrange

Arrange sorts rows by given column(s).

```{r}
flights %>% arrange(dep_delay)
```

## Arrange in descending order

```{r}
flights %>% arrange(desc(dep_delay))
```

## Arrange with multiple keys

```{r}
flights %>% arrange(dep_delay, arr_delay)
```


## **$\Rightarrow$** Arrange exercises

1. Sort the rows by the amount of time made up during the flight, from most to least.

2. Four of the flights in our Excel file (rows 840-843) had missing values for `dep_delay`. How does R represent missing values? When the rows are sorted by `dep_delay`, where do these end up?

## Select

Select columns you would like to see.

```{r}
flights %>% select(air_time, dep_delay, arr_delay)
```

## Move some columns to front

Use the `everything()` helper to select columns you want on the left.

```{r}
flights %>% select(carrier, flight, everything())
```

## Remove selected columns

```{r}
flights %>% select(-year, -month)
```

## **$\Rightarrow$** Select exercises

1. Make the first column the `tailnum`, and remove the `time_hour` column.

2. Try using `select(contains("time"))`. What happens?

# *Pipes* {data-background="https://static.cms.yp.ca/ecms/media/1/pipes-1445309646-600x360.jpg"}

## Combinations with pipes

- Often we want to perform operations in sequence.
- The "pipe" operator `%>%` sends the result of the last command to a new command.
- For readability, do one operation per line. Work from a code block, not the console.

## Filter then arrange

```{r}
flights %>% 
  filter(arr_delay > 100) %>%
  arrange(arr_delay)
```

## Filter then arrange then select

```{r}
flights %>% 
  filter(arr_delay > 100) %>%
  arrange(arr_delay) %>%
  select(carrier, flight, arr_delay)
```

## **$\Rightarrow$** Combo exercises

1. For flights departing from `JFK`, sort them by distance (largest to smallest), and display the carrier and destination city.

2. Does the order of `arrange` and `select` matter? Experiment.

# *Computations* {data-background="https://image1.masterfile.com/getImage/NzAwLTAyMjAwNTc4ZW4uMDAwMDAwMDA=AM5J$Z/700-02200578en_Masterfile.jpg"}

## Mutate

Add new columns based on existing ones. The new column goes at the end.

```{r}
flights %>% mutate(gain = dep_delay - arr_delay)
```

## Mutate and select

Let's select some of the columns so we can see our new variables.

```{r}
flights %>% 
  mutate(gain = dep_delay - arr_delay) %>%
  select(tailnum, origin, dest, gain)
```

## Multiple new columns

```{r}
flights %>% 
  mutate(gain = dep_delay - arr_delay,
         hours = air_time / 60,
         gain_per_hour = gain / hours) %>%
  select(tailnum, origin, dest, gain, hours, gain_per_hour)
```

## Transmute

`transmute` just gives you back the new variables.

```{r}
flights %>% 
  transmute(gain = dep_delay - arr_delay,
         hours = air_time / 60,
         gain_per_hour = gain / hours) 
```

## **$\Rightarrow$** Mutate exercises

1. Compare `air_time` with `arr_time - dep_time`. What do you expect to see? What do you see? What needs to be fixed?

2. Sort the flights by `gain_per_hour` (see above) and display the 20 highest gains, in order, along with the carrier and flight number.

## Summarize

Collapse the data into a single row.

```{r}
flights %>% 
  summarize(mean_dist = mean(distance), 
            sd_dist = sd(distance))
```

## Grouped summaries



## **$\Rightarrow$** Summarize exercises

