---
title: "Dealing with Data and Packages"
date: "May 7, 2019"
output: 
  revealjs::revealjs_presentation:
    fig_width: 14
    fig_height: 7
    self_contained: true
    theme: black
    highlight: zenburn
    css: slidesdjh.css
    center: false
    transition: slide
    reveal_options:
      controls: true
      progress: false
      width: 1080
      height: 600
---

# Extending R with packages {data-background="https://pbs.twimg.com/media/CvzEQcfWIAAIs-N.jpg" data-background-size="contain"}

## Packages in R

>- So far, we have been using "base" R.
>- One of the most compelling reasons to use R is the enormous ecosystem of packages that extend its capabilities.
>- [CRAN task view](https://cran.rstudio.com/web/views/)

## Install the Tidyverse

We will be using a suite of packages designed for data science.

**$\Rightarrow$** Install the `tidyverse` by executing the following command in the console:

```{r, eval=FALSE}
install.packages("tidyverse")
```

This installation may take a few moments.

# *Getting Data into R* {data-background="https://tr2.cbsistatic.com/hub/i/2016/05/27/b6dd6d05-84b0-4eeb-ad7f-ddea46d0f15c/unexpected-error.jpg" data-background-size="contain"}

## Reproducibilty principles

>- We want other people to be able to look at our data and reproduce our statistical analyses.
>- We should make our data public.
>- We should make our code public.
>- Importing data into a statistical package often involves choices. These choices should be reproducible.
>- Therefore, **our code should import our data.** Avoid using menus or mouse clicks to import data.

## Importing Data from a website

- In the [last lesson](intro_stats.html), we read most of our data sets off of websites.
- For many R commands, `file` can be a path to a local file, or a URL.
- This is fairly reproducible. (It would be better if the code and data were in the same place.)

## **$\Rightarrow$** Make a new project

1. In the *File* menu, choose *New Project...*. In the dialog that appears, select *New Directory* followed by  *New Project*.

2. For the *Directory Name*, use `DataWorkshop`. Click on *Browse* and find a subdirectory where you want this new directory to reside (e.g., Documents, Dropbox, etc.). You will need to be able to find this directory later.

3. Click on *Create Project*. A new empty project should load. A file named `DataWorkshop.Rproj` should appear in the *Files* tab in the lower right pane.

## **$\Rightarrow$** Make a new RMarkdown document

1. In the *File* menu, choose *New File > RMarkdown...*.
2. In the "New R Markdown" dialog box, give the document a title. Keep the other default settings (Document, HTML).
3. Clicking OK will open a new RMarkdown document with some boilerplate code. *Delete* everything below line 7. 
4. *Save* the document as `datatidy.Rmd` in the project directory `DataWorkshop`. After you save the file, it should appear in the *Files* tab, along with the file `DataWorkshop.Rproj`.

## Importing Data from a local file

- Quite often, we have data in a local file we want to use.
- We should put our data where our code is: the project folder.

## **$\Rightarrow$** Download some data

1. Download the Excel file [flights.xlsx](http://math.westmont.edu/data/flights.xlsx) to your computer. **Make sure it ends up in the project folder `DataWorkshop` that you just created.** Depending on your browser settings, it may end up in your `Downloads` folder, and you will have to move it to the `DataWorkshop` folder. 
2. Refresh the view in the *Files* pane of your project, and make sure that the `flights.xlsx` file is there.

## **$\Rightarrow$** Read the Excel file and view 3 ways

1. Execute the following command in the terminal. The first line loads the tidyverse package `readxl`, and the second line reads the entire spreadsheet into the `flights` variable.

```{r}
library(readxl)
flights <- read_excel("flights.xlsx")
```

2. Type `flights` in the terminal.

3. In the *Environment* tab in the upper-right corner, click on the triangle next to the word `flights`.
4. In the *Environment* tab in the upper-right corner, click on the word `flights`.

## Tibbles

- The variable `flights` is a `tibble`, which is a data frame with additional features.
- Tibbles are part of the tidyverse.
- Tibbles can be treated just like data frames

# *Shaping Data* {data-background="https://laughingsquid.com/wp-content/uploads/2017/10/game-of-thrones-timelapse-of-an-artist-sculpting-and-airbrushing-a-bust-of-the-night-king.gif" data-background-size="contain"}

## Shaping your data with dplyr

The remainder of this lesson is adapted from the [Data transformation](https://r4ds.had.co.nz/transform.html) chapter in *R for Data Science*.

## Dplyr basics: Five functions

- Pick observations by their values (`filter()`).
- Reorder the rows (`arrange()`).
- Pick variables by their names (`select()`).
- Create new variables with functions of existing variables (`mutate()`).
- Collapse many values down to a single summary (`summarise()`).

**$\Rightarrow$** Load the `dplyr` package in the console.

```{r}
library(dplyr)
```


## Filter

Select observations based on their values.

```{r}
flights %>% filter(arr_delay > 100)
```




