---
title: "Getting Started using R"
date: "May 6, 2019"
output: 
  revealjs::revealjs_presentation:
    fig_width: 14
    fig_height: 7
    self_contained: true
    theme: black
    highlight: zenburn
    css: slidesdjh.css
    center: false
    transition: slide
    reveal_options:
      controls: true
      progress: false
      width: 1080
      height: 600
---

# Getting Started with R

## Why R?

>- Reproducibility
>- Powerful
>- Free (as in speech)
>- Free (as in beer)

## Overview

>- Commands in the console
>- Vectors
>- Data Frames
>- Workflow
>     - Scripts
>     - RMarkdown notebooks

## Workshop materials

Everything is posted at: https://djhunter.github.io/Rworkshop

Slides can be viewed on any web browser. Keyboard shortcuts:

| shortcut | result |
| --- | --- |
| o | Toggle Overview mode (see all the slides) |
| f | Enter fullscreen mode | 
| ESC | Exit fullscreen mode |
| space | next slide |
| arrows | navigate up/down/left/right |
| ? | See other shortcuts |

The **$\Rightarrow$** symbol indicates something for you to do.

# RStudio Interface

## {data-background="https://info201.github.io/img/r-intro/rstudio-interface.png" data-background-size="contain"}

# The Console

## Use R Interactively 

In the console window, you can type commands and get answers. R understands basic arithmetic.

```{r}
2+2
```

**$\Rightarrow$** Try some basic arithmetic commands in the console window. 

## Vectors

- Often you want to deal with several numbers at once.
- The `c( )` operator *combines* its contents into a *vector*.

```{r}
c(2,4,6,2,10)
c(2,4,6,2,10) + c(10, 20, 30, 40, 50)
c(2,4,6,2,10) * 1000
```

## Variables

- Often you want to save a vector so you can reuse it.
- The assignment operator `<-` saves a variable.

```{r}
joe <- c(2,4,6,2,10)
joe^2
sum(joe)
mean(joe)
```

## Practice with variables

**$\Rightarrow$** Use the console to compute 
$$\frac{1}{2^2} + \frac{1}{3^2} + \frac{1}{4^2} + \frac{1}{5^2} +\frac{1}{6^2} +\frac{1}{7^2} +\frac{1}{8^2} +\frac{1}{9^2}  + \frac{1}{10^2}.$$ 
Start by assigning `c(2,3,4,5,6,7,8,9,10)` to a variable with a name you make up.

# Workflow

## Saving your work

Two ways to save your work:

>- R Scripts (`foo.R` files)
>     - Simple, universal
>- RMarkdown documents (`foo.Rmd` files)
>     - Combinations of scripts and other text
>     - Can be exported (HTML, slides, PDF, etc)
>- Workflow 
>     - Experiment in console
>     - Save what worked as R Scripts or RMarkdown documents

## {data-background="https://info201.github.io/img/r-intro/rstudio-interface.png" data-background-size="contain"}

## Make an R Script

**$\Rightarrow$** Follow the steps below to save your work from the last exercise.

1. Click on the *History* tab in the upper-right pane. Locate the lines of code that successfully computed the sum $\frac{1}{2^2} + \frac{1}{3^2} + \cdots + \frac{1}{10^2}$. 
2. In the *File* menu, choose *New File > R Script*.
3. Click on the lines you located in #1, and click the *$\Leftarrow$ To Source* button. 
4. Save your script to a file with a location and name you will remember. The file name should end in `.R`.
5. Test your script by clicking the *$\Rightarrow$  Source* button in the upper-left pane. The contents should execute in the console.

## Update your script

You can define a sequence of numbers using the `:` operator.

```{r}
2:10
```

**$\Rightarrow$** Update your script so it computes $\frac{1}{2^2} + \frac{1}{3^2} + \cdots + \frac{1}{1000000^2}$. (Compare the answer to $\pi^2/6-1$.)

## Make an RMarkdown Document


# Variables and Data Frames

