---
title: "Graphics"
date: "May 7, 2019"
output: 
  revealjs::revealjs_presentation:
    fig_width: 14
    fig_height: 7
    self_contained: true
    theme: black
    highlight: zenburn
    css: slidesdjh.css
    center: false
    transition: slide
    reveal_options:
      controls: true
      progress: false
      width: 1080
      height: 600
  html_document:
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: true
    df_print: paged
    theme: spacelab
---

# *Better Graphics with ggplot* {data-background="http://4.bp.blogspot.com/-s-Qbt193kz4/UX8GoEuZo4I/AAAAAAAAGXI/ZTDOEYyq4ak/s1600/DSC04879.JPG" data-background-size="contain"}

## Grammar of Graphics

>- So far we have been using the "base" graphics of R. 
>- Base graphics can look good, but it takes some work.
>- In this lesson, we'll explore the `ggplot2` package, which is part of the tidyverse.
>- `ggplot2` graphics look better "out of the box", and the syntax follows the tidyverse philosopy.

## **$\Rightarrow$** Make a new RMarkdown document

1. Make sure your are in your `DataWorkshop` project. If not, switch to it.
2. In the *File* menu, choose *New File > RMarkdown...* and title a new document.
3. Clicking OK will open a new RMarkdown document with some boilerplate code. *Delete* everything below line 7. 
4. *Save* the document as `tidyplots.Rmd` in the project directory `DataWorkshop`. After you save the file, it should appear in the *Files* tab, along with the file `DataWorkshop.Rproj`.

## **$\Rightarrow$** Load the tidyverse

```{r}
library(tidyverse)
```

## **$\Rightarrow$**  look at `mpg` tibble

The `tidyverse` comes with a built-in tibble called `mpg`. To make it appear in the upper-right pane, type `View(mpg)` in the console. You can also just echo the name:

```{r}
mpg
```

## **$\Rightarrow$** Review Exercise

Use the tidyverse data transformations to create a list of all the auto manufacturers, along with the average city mpg of their vehicles, sorted from most fuel-efficient to least.

# using GGplot

## Data Visualization

Most of this lesson is adapted from [Chapter 3](https://r4ds.had.co.nz/data-visualisation.html) of *R for Data Science*.

## ggplot grammar

- `ggplot()` creates a blank graph.
- `ggplot(mpg)` associates the `mpg` data.
- `ggplot(mpg) + geom_point()` adds a scatterplot to the graph, but we need to specify what variables to use.
-  `aes()` specifies which variables are represented by different properties of the graph (aesthetics).
- There are lots of different `geom`'s. You can add layers to the plot with `+`'s.

## ggplot scatterplot

```{r, fig.width=7, fig.height=4, fig.align='center'}
ggplot(mpg) + geom_point(aes(x=displ, y=hwy))
```

## Coloring factors

```{r, fig.width=7, fig.height=4, fig.align='center'}
ggplot(mpg) + geom_point(aes(x=displ, y=hwy, color = class))
```


