---
title: "Graphics"
date: "May 7, 2019"
output: 
  revealjs::revealjs_presentation:
    fig_width: 14
    fig_height: 7
    self_contained: true
    theme: black
    highlight: zenburn
    css: slidesdjh.css
    center: false
    transition: slide
    reveal_options:
      controls: true
      progress: false
      width: 1080
      height: 600
  html_document:
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: true
    df_print: paged
    theme: spacelab
---

# *Better Graphics with ggplot* {data-background="http://4.bp.blogspot.com/-s-Qbt193kz4/UX8GoEuZo4I/AAAAAAAAGXI/ZTDOEYyq4ak/s1600/DSC04879.JPG" data-background-size="contain"}

## Grammar of Graphics

>- So far we have been using the "base" graphics of R. 
>- Base graphics can look good, but it takes some work.
>- In this lesson, we'll explore the `ggplot2` package, which is part of the tidyverse.
>- `ggplot2` graphics look better "out of the box", and the syntax follows the tidyverse philosopy.

## **$\Rightarrow$** Make a new RMarkdown document

1. Make sure you are in your `DataWorkshop` project. If not, switch to it.
2. In the *File* menu, choose *New File > RMarkdown...* and title a new document.
3. Clicking OK will open a new RMarkdown document with some boilerplate code. *Delete* everything below line 7. 
4. *Save* the document as `tidyplots.Rmd` in the project directory `DataWorkshop`. After you save the file, it should appear in the *Files* tab, along with the file `DataWorkshop.Rproj`.

## **$\Rightarrow$** Load the tidyverse

```{r}
library(tidyverse)
```

## **$\Rightarrow$**  look at `mpg` tibble

The `tidyverse` comes with a built-in tibble called `mpg`. To make it appear in the upper-right pane, type `View(mpg)` in the console. You can also just echo the name:

```{r}
mpg
```

## **$\Rightarrow$** Review Exercise

Use the tidyverse data transformations to create a list of all the auto manufacturers, along with the average city mpg of their vehicles, sorted from most fuel-efficient to least.

# *using GGplot* {data-background="https://www.r-graph-gallery.com/wp-content/uploads/2017/12/327_Chloropleth_geoJSON_ggplot2_4-800x600.png"}

## Data Visualization

Most of this lesson is adapted from [Chapter 3](https://r4ds.had.co.nz/data-visualisation.html) of *R for Data Science*.

## ggplot grammar

- `ggplot()` creates a blank graph.
- `ggplot(mpg)` associates the `mpg` data.
- `ggplot(mpg) + geom_point()` adds a scatterplot to the graph, but we need to specify what variables to use.
-  `aes()` specifies which variables are represented by different properties of the graph (aesthetics).
- There are lots of different `geom`'s. You can add layers to the plot with `+`'s.

# *Scatterplots* {data-background="https://cdn1.autoexpress.co.uk/sites/autoexpressuk/files/2016/07/dsc_6125_0.jpg" data-background-size="contain"}

## ggplot scatterplot

```{r, fig.width=7, fig.height=4, fig.align='center'}
ggplot(mpg) + geom_point(aes(x=displ, y=hwy))
```

## Coloring factors

```{r, fig.width=7, fig.height=4, fig.align='center'}
ggplot(mpg) + geom_point(aes(x=displ, y=hwy, color = class))
```

## Sizing dots

```{r, fig.width=7, fig.height=4, fig.align='center'}
ggplot(mpg) + geom_point(aes(x=displ, y=hwy, color = class, size=cyl))
```

## Manual settings

```{r, fig.width=7, fig.height=4, fig.align='center'}
ggplot(mpg) + geom_point(aes(x=displ, y=hwy), color = "blue")
```

## Manual settings (wrong: why?)

```{r, fig.width=7, fig.height=4, fig.align='center'}
ggplot(mpg) + geom_point(aes(x=displ, y=hwy, color = "blue"))
```

## **$\Rightarrow$** Scatterplot exercises

1. Map a continuous variable to `color`, `size`, and `shape`. 

2. Map the same variable to multiple aesthetics.

3. What does the `stroke` aesthetic do? What shapes does it work with? (Hint: use `vignette("ggplot2-specs")`.)

4. What happens if you map an aesthetic to something other than a variable name, like `aes(color = displ < 5)`? Note, you’ll also need to specify `x` and `y`.

# *Adding geometries* {data-background="http://oscarperpinan.github.io/spacetime-vis/images/unemployUSAThemeRiver.png"}

## Facet Grid

```{r, fig.width=7, fig.height=4, fig.align='center'}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_grid(drv ~ cyl)
```

## Facet Grid (collapsed)

```{r, fig.width=7, fig.height=4, fig.align='center'}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_grid(. ~ cyl)
```

## Facet Wrap

```{r, fig.width=7, fig.height=4, fig.align='center'}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_wrap(~ class, nrow = 2)
```

## ggplot2 geometries

**$\Rightarrow$** In the lower-right pane, click on the *Packages* tab, then find the link for `ggplot2`. Clicking this link should bring up the package help pages; scroll down to the "G" section and observe all of the available geometries.

## Recall: `geom_point`

```{r, fig.width=7, fig.height=4, fig.align='center'}
ggplot(mpg) + 
  geom_point(aes(x = displ, y = hwy))
```

## Replace `geom_point` with `geom_smooth`

```{r, fig.width=7, fig.height=4, fig.align='center', message=FALSE}
ggplot(mpg) + 
  geom_smooth(aes(x = displ, y = hwy))
```

## Combine geometries with +

```{r, fig.width=7, fig.height=4, fig.align='center', message=FALSE}
ggplot(mpg) + 
  geom_point(aes(x = displ, y = hwy)) +
  geom_smooth(aes(x = displ, y = hwy))
```

## Change the `method`

```{r, fig.width=7, fig.height=4, fig.align='center', message=FALSE}
ggplot(mpg) + 
  geom_point(aes(x = displ, y = hwy)) +
  geom_smooth(aes(x = displ, y = hwy), method = "lm")
```

## `linetype` aesthetic

```{r, fig.width=7, fig.height=4, fig.align='center', message=FALSE}
ggplot(mpg) + 
  geom_smooth(aes(x = displ, y = hwy, linetype = drv))
```

## **$\Rightarrow$** Geometry exercises

1. Try to produce the following plot:

```{r, fig.width=7, fig.height=4, fig.align='center', message=FALSE, echo=FALSE}
ggplot(mpg) + 
  geom_smooth(aes(x = displ, y = hwy, linetype = drv, color=drv)) +
  geom_point(aes(x = displ, y = hwy, color = drv)) 
```

## **$\Rightarrow$** Geometry exercises

2. Try to produce the following plot:

```{r, fig.width=7, fig.height=4, fig.align='center', message=FALSE, echo=FALSE}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point(mapping = aes(color = class)) + 
  geom_smooth() +
  geom_smooth(method = "lm", color="purple")
```

## **$\Rightarrow$** Geometry exercises

3. Read `?facet_wrap`. What does `nrow` do? What does `ncol` do? Why doesn’t `facet_grid()` have these arguments?

4. Investigate the `ggplot2` geometry functions (click on the `ggplot2` link in the packages tab). What `geom` would you use to draw a line chart? A boxplot? A histogram? An area chart?

# *Bar graphs and more* {data-background="http://www.colonialjewelers.com/wp-content/uploads/2017/01/diamonds.jpg"}

## **$\Rightarrow$** Explore the `diamonds` tibble

Try `View(diamonds)` in the console to see the contents of this built-in tidyverse data set.

```{r}
diamonds
```

## Bar charts

```{r, fig.width=7, fig.height=4, fig.align='center'}
ggplot(diamonds) + 
  geom_bar(aes(x = cut))
```

## Bar charts (segmented)

```{r, fig.width=7, fig.height=4, fig.align='center'}
ggplot(diamonds) + 
  geom_bar(aes(x = cut, fill=clarity))
```

```{r, fig.width=7, fig.height=4, fig.align='center'}
ggplot(diamonds) + 
  geom_bar(aes(x = cut, fill=clarity), position="fill")
```

## Bar charts (grouped)

```{r, fig.width=7, fig.height=4, fig.align='center'}
ggplot(diamonds) + 
  geom_bar(aes(x = cut, fill=clarity), position="dodge")
```

## Boxplots

```{r, fig.width=7, fig.height=4, fig.align='center'}
ggplot(diamonds) + 
  geom_boxplot(aes(x = cut, y = price))
```

## Using pipes

```{r, fig.width=7, fig.height=4, fig.align='center', eval=FALSE}
ggplot(diamonds) + 
  geom_boxplot(aes(x = cut, y = price))
```

does the same thing as:

```{r, fig.width=7, fig.height=4, fig.align='center', eval=FALSE}
diamonds %>% ggplot() + 
  geom_boxplot(aes(x = cut, y = price))
```

## Filter, then pipe to ggplot

```{r, fig.width=7, fig.height=4, fig.align='center'}
diamonds %>% filter(price<7500) %>% ggplot() + 
  geom_boxplot(aes(x = cut, y = price))
```

## Density plots

```{r, fig.width=7, fig.height=4, fig.align='center'}
ggplot(diamonds) + 
  geom_density(aes(x = price, fill=cut))
```

## Density plots (with transparency)

```{r, fig.width=7, fig.height=4, fig.align='center'}
ggplot(diamonds) + 
  geom_density(aes(x = price, fill=cut), alpha=0.3)
```

## **$\Rightarrow$** More exercises

1. Create a density plot for diamonds priced between 2500 and 7500, grouped by cut. Then make a histogram of the same thing.

```{r, fig.width=7, fig.height=4, fig.align='center', eval=FALSE, echo=FALSE}
diamonds %>% filter(price > 2500, price < 7500) %>% ggplot() + 
  geom_density(aes(x = price, fill=cut), alpha=0.3)
diamonds %>% filter(price > 2500, price < 7500) %>% ggplot() + 
  geom_histogram(aes(x = price, fill=cut), alpha=0.3)
```

2. Try appending `+ coord_flip()` to the end of a ggplot sum. 

3. Try appending `+ coord_polar()` to the end of a ggplot sum for some of the above boxplots. Can you figure out how to make a traditional pie chart? ([Not that you ever should](https://www.businessinsider.com/pie-charts-are-the-worst-2013-6).)

```{r, fig.width=7, fig.height=4, fig.align='center', eval=FALSE, echo=FALSE}
ggplot(diamonds) + 
  geom_bar(aes(x = "", fill=clarity), position="fill") + coord_polar("y")
```

4. Try appending `+ theme_classic()`. Investigate other `ggplot2` themes.

```{r, fig.width=7, fig.height=4, fig.align='center', eval=FALSE, echo=FALSE}
ggplot(diamonds) + 
  geom_density(aes(x = price, fill=cut)) + theme_classic()
```

5. Using the `mpg` data, make an appropriate chart showing the average city mpg of each manufacturer, in order.

```{r, fig.width=7, fig.height=4, fig.align='center', eval=FALSE, echo=FALSE}
mpg %>% 
  group_by(manufacturer) %>%
  summarize(mpg = mean(cty)) %>%
  arrange(mpg) %>%
  mutate(row = row_number()) %>%
  ggplot() + 
  geom_bar(aes(x=reorder(manufacturer, row), 
               y=mpg, 
               fill=manufacturer), 
           stat = "identity", 
           show.legend=FALSE) + 
  coord_flip()
```


