---
title: "Introductory Statistics in 5 Minutes"
date: "May 6, 2019"
output: 
  revealjs::revealjs_presentation:
    fig_width: 14
    fig_height: 7
    self_contained: true
    theme: black
    highlight: zenburn
    css: slidesdjh.css
    center: false
    transition: slide
    reveal_options:
      controls: true
      progress: false
      width: 1080
      height: 600
---

# *MA-5 in 5 Minutes* {data-background="http://pfeifferpartners.com/wp-content/uploads/2016/03/pfeiffer-westmont-int-winter-main-lobby-stair-M.jpg" data-background-size="contain"}

## Basic Statistical Tests in R

>- Once the data is loaded, most tests are one line of code.
>- Simple graphs are also one-liners (base graphics).

## MA-5 Topics I cover

- One proportion and one mean
- Two proportions and two means (independent samples and paired)
- Multiple proportions ($\chi^2$)
- Multiple means (ANOVA)
- Regression and Correlation
- Multiple linear regression
- MANOVA, Factorial ANOVA, Logistic Regression (optional)

>- *Nonparametric Resampling Techniques for all of the above*

## Introduction to Statistical Investigations

- Data and Simulation-Based Applets: http://www.isi-stats.com/isi/applets.html
- Slides from last semester: http://math.westmont.edu/ma5

# *Keep track of notes in RMarkdown* {data-background="https://i2.wp.com/mimshouse.com/wp-content/uploads/16213476783_e5bb974e16_k.jpg" data-background-size="contain"}

## **$\Rightarrow$** Make an RMarkdown Document

Let's keep track of our work in this lesson using RMarkdown.

1. In the *File* menu, choose *New File > RMarkdown...*.
2. In the "New R Markdown" dialog box, give the document a title. Keep the other default settings (Document, HTML).
3. Clicking OK will open a new RMarkdown document with some boilerplate code. *Delete* everything below line 7. 
4. *Save* the file to a location and name you will remember. The file name should end in `.Rmd`. (e.g., `introstats.Rmd`)

# *One Sample Tests* {data-background="https://cdn.cnn.com/cnnnext/dam/assets/151119194429-04-cancer-dogs-super-169.jpg" data-background-size="contain"}

## One-Proportion Test

$$
H_0: \pi = 0.5 \\ 
H_a: \pi \neq 0.5 \\ 
\hat{p} = 54/83, \quad n=83
$$

```{r, eval=FALSE}
prop.test(54, 83, p = 0.5)
```

## One-Proportion Test

```{r}
prop.test(54, 83, p = 0.5)
```

## **$\Rightarrow$** Explore the help menus

1. In the console, type `?prop.test`. Read the help menu that appears.

2. Test $H_0: \pi = 0.5$ versus $H_a: \pi > 0.5$ (*one-sided* alternative) for $\hat{p} = 54/83$, $n=83$, *without* the continuity correction.

3. Read about the exact binomial test: `?binom.test`.

4. Repeat the test in #2 using the exact binomial test.

## **$\Rightarrow$** Save your work

5. Above the editor pane for your RMarkdown document, click on the *$^+C$ Insert* button, and choose *R* from the drop down menu. Inside the code chunk that appears, enter the code that workd for steps 1-4. 
6. Add some text above or below the code chunk explaining what these commands do. 
7. *Knit* and view the HTML output.

## One-Sample $t$-test

sleep 
```{r}
sleep <- read.table("http://math.westmont.edu/ma5/classSleep.txt", header = TRUE)
sleep
```

## One-Sample $t$-test

$$
H_0: \mu = 0 \\
H_a: \mu \neq 0
$$

```{r, eval=FALSE}
t.test(sleep)
```

## One-Sample $t$-test

```{r}
t.test(sleep)
```

## **$\Rightarrow$** Customize the $t$-test

1. Load the data using the following command. (Copy and paste it into the console.)

```{r, eval=FALSE}
sleep <- read.table("http://math.westmont.edu/ma5/classSleep.txt", header = TRUE)
```

2. See the help page using `?t.test` in the console. 

3. Change the *alternative* hypothesis to $H_a : \mu < 8$ and generate a p-value. Also, generate a *99%* confidence interval.

4. Save your code and make some notes about it in your RMarkdown document.

# *Explanatory/Response Scenarios* {data-background="https://pages.wustl.edu/files/pages/olsen/site/slideshow/img3260.jpg" data-background-size="contain"}

## Explanatory/Response variables

| Explanatory | Response | Scenario/Test |
| ---- | ---- | ---- |
| Categorical (2 levels) | Categorical | two-proportion $z$-test |
| Categorical (2 levels, independent) | Quantitative | two-sample $t$-test |
| Categorical (2 levels, paired) | Quantitative | paired $t$-test |
| Categorical (>2 levels) | Categorical | $\chi^2$ test |
| Categorical (>2 levels) | Quantitative | ANOVA $F$-test |
| Quantitative | Quantitative | Regression/Correlation |

## Two Proportion $z$-test

```{r}
gilbert <- read.table("http://www.isi-stats.com/isi/data/chap5/Gilbert.txt", header = TRUE)
head(gilbert) # just show the first few lines
```

## Two Proportion $z$-test

```{r}
table(gilbert)
```

## Two Proportion $z$-test

There were 34 deaths on the 1384 shifts when Gilbert didn't work, and 40 deaths on the 257 shifts when she did work.

$$
H_0: \pi_1 - \pi_2 = 0 \\
H_a: \pi_1 - \pi_2 \neq 0 
$$

```{r, eval=FALSE}
prop.test(c(34, 40), c(1384, 257))
```

## Two Proportion $z$-test

There were 34 deaths on the 1384 shifts when Gilbert didn't work, and 40 deaths on the 257 shifts when she did work.

```{r}
prop.test(c(34, 40), c(1384, 257))
```

## Two Proportion $z$-test

Instead of giving the counts, you can just send `prop.test` the table:

```{r}
prop.test(table(gilbert))
```

# *Scenarios using formulas* {data-background="https://www.open.edu/openlearn/ocw/pluginfile.php/775971/mod_resource/content/0/mu120_4m6_cover_image_1.jpg"}

## Formulas in R

- Explanatory (x) and response (y) variables are usually denoted using "tilde" notation: `y ~ x`
- Works in different contexts (tests, plots)
- Can also have multiple response, explanatory variables

## Formulas in plots

```{r, fig.align='center', fig.height=4, fig.width=4}
plot(Patient ~ GilbertWorked., data = gilbert)
```

## Two-sample $t$-test

```{r}
friends <- read.table("http://www.isi-stats.com/isi/data/chap6/CloseFriends.txt", header = TRUE)
summary(friends)
```

## Two-sample $t$-test

```{r, fig.align='center', fig.height=4, fig.width=4}
plot(Friends ~ Sex, data = friends)
```

## Two-sample $t$-test

$$
H_0: \mu_1 - \mu_2 = 0 \\
H_a: \mu_1 - \mu_2 \neq 0 
$$

```{r, eval=FALSE}
t.test(Friends ~ Sex, data = friends)
```

## Two-sample $t$-test

```{r}
t.test(Friends ~ Sex, data = friends)
```

## Paired $t$-test

```{r}
rounding1st <- read.table("http://www.isi-stats.com/isi/data/chap7/FirstBase.txt", header=TRUE)
summary(rounding1st)
```

## Paired $t$-test

$$
H_0: \mu_d  = 0 \\
H_a: \mu_d  \neq 0 
$$

```{r, eval = FALSE}
t.test(rounding1st$narrow, rounding1st$wide, paired = TRUE)
```

## Paired $t$-test

```{r}
t.test(rounding1st$narrow, rounding1st$wide, paired = TRUE)
```

## Chi-square test

```{r}
apunc <- read.table("http://www.isi-stats.com/isi/data/chap8/Acupuncture.txt", header = TRUE)
table(apunc)
head(apunc)
```

## Chi-square test

```{r, fig.align='center', fig.height=4, fig.width=5}
plot(Improvement ~ Acupunture, data = apunc)
```

## Chi-square test

```{r}
chisq.test(table(apunc))
```

## One-way ANOVA

```{r, fig.align='center', fig.height=4, fig.width=5}
diets <- read.table("http://www.isi-stats.com/isi/data/chap9/Diets.txt", header = TRUE)
plot(BMI ~ Diet, data = diets)
```

## One-way ANOVA

```{r}
d.aov <- aov(BMI ~ Diet, data = diets)
anova(d.aov)
```

## Tukey HSD follow-up

```{r}
TukeyHSD(d.aov)
```

## Regression/Corellation

```{r, fig.align='center', fig.height=4, fig.width=5}
textbooks <- read.table("http://www.isi-stats.com/isi/data/chap10/TextbookPrices.txt", header = TRUE)
plot(Price ~ Pages, data = textbooks)
```

## Regression/Corellation

```{r}
summary(lm(Price ~ Pages, data = textbooks))
```

## Correlation Matrix

```{r}
cor(textbooks)
```

# *Resources for other scenarios* {data-background="https://imgs.xkcd.com/comics/correlation.png" data-background-size="contain"}

## UCLA Test Chooser

https://stats.idre.ucla.edu/other/mult-pkg/whatstat/

## **$\Rightarrow$** Breakfast Cereal

**$\Rightarrow$** Do the following in the console, and transfer to your RMarkdown document, or work directly from the RMarkdown document.

1. Load the breakfast cereal data:

```{r}
cereal <- read.table("http://math.westmont.edu/data/cereal.csv", header=TRUE, sep=";")
```

2. This data set is quite large, so view it by clicking on the *Environment* tab in the upper-right pane, and click on the `cereal` name to see the data as a spreadsheet.

3. Produce a [correlation matrix](#/correlation-matrix) using [columns](getting_started.html#/selecting-noncontiguous-columns) 7, 8, 9, 10, and 16.

## **$\Rightarrow$** Regression

4. Make a [scatterplot](#/regressioncorellation) of `rating` versus `sugar`.

5. Produce a [regression table](#/regressioncorellation-1) for `rating` as explained by `sugar`.

6. Produce a regression table for `rating` as explained by `fiber`.

7. Produce a regression table for `rating` as explained by `sugar + fiber`. (That's multiple regression.)

## **$\Rightarrow$** ANOVA

8. We'd like to treat `shelf` as a categorical variable (factor). Do so:

```{r}
cereal$shelf <- factor(cereal$shelf) # make R treat shelf as categorical
```

9. Produce an [ANOVA table](#/one-way-anova-1) for `sugar` as explained by `shelf`. Do a [post-hoc test](#/tukey-hsd-follow-up), too.

10. Produce an ANOVA table for `sugar` as explained by `shelf * mfr`. (That's factorial ANOVA.) Do a post-hoc test, too. 

**$\Rightarrow$** If you have been working from the console, save all your code into your RMarkdown document.

# *Tricked-out RMarkdown* {data-background="https://i.pinimg.com/originals/94/53/62/945362d91788247548930f28520b9a03.gif" data-background-size="contain"}

## RMarkdown formatting

RMarkdown supports lots of formatting options. See:
https://bookdown.org/yihui/rmarkdown/markdown-syntax.html#block-level-elements

**$\Rightarrow$** Experiment with adding some formattting to your RMarkdown document. Consider adding:

- Headers of different levels
- A bulleted list
- A blockquote

You will have to *Knit* to see what worked.

## Tricked-out HTML

There are lots of HTML options. See: https://bookdown.org/yihui/rmarkdown/html-document.html

**$\Rightarrow$** Experiment with some HTML optons. Try:

- Adding a table of contents.
- A different document theme.





