---
title: "RMarkdown Worksheets"
author: "David J. Hunter"
output:
  html_document:
    toc: true
    number_sections: false
    toc_float:
      collapsed: false
      smooth_scroll: true
    df_print: paged
    theme: spacelab
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo=FALSE}
library(metathis)
meta() %>%
  meta_description(
    "A collection of self-guided RMarkdown activities to get started using R and RStudio."
  ) %>%
  meta_name("github-repo" = "djhunter/Rworkshop") %>%
  meta_viewport() %>%
  meta_social(
    title = "RMarkdown Worksheets",
    url = "https://djhunter.github.io/Rworkshop/worksheets/index.html",
    image = "https://d33wubrfki0l68.cloudfront.net/521a038ed009b97bf73eb0a653b1cb7e66645231/8e3fd/assets/img/rstudio-icon.png",
    image_alt = "RStudio Logo",
    og_author = c("David J. Hunter")
  )
```

When downloading these worksheets, it may be necessary to right-click (or two-finger-click) and choose *Save Link As...*.

# Getting Started

## [rmarkdown_workflow.Rmd](rmarkdown_workflow.Rmd)

For a self-guided tour of the RStudio interface, open this file in RStudio and read through it, doing the tasks indicated in **bold**.

## [simple.Rmd](simple.Rmd) 

This barebones RMarkdown template illustrates how to *Knit* and use code blocks. 

## [report.Rmd](report.Rmd)

Insert some simple commands and email yourself a report.

# Introductory Statistics

# Dealing with Data

# Graphics

[Back to Workshop Page](../index.html)