---
title: "Biology Department introduction to R"
output:
  pdf_document:
    includes:
      in_header: notesheader.tex
    keep_tex: yes
  html_document:
    df_print: paged
  revealjs::revealjs_presentation:
    center: no
    css: slidesdjh.css
    fig_height: 7
    fig_width: 14
    highlight: zenburn
    reveal_options:
      controls: yes
      height: 600
      progress: no
      width: 1080
    self_contained: yes
    theme: black
    transition: slide
date: "January 28, 2020"
---

# *Getting Started with R* {data-background="https://upload.wikimedia.org/wikipedia/commons/thumb/1/1b/R_logo.svg/724px-R_logo.svg.png" data-background-size="contain"}

## Why R?

- Reproducibility
- Powerful
- Very deep community of support
- Free (*gratis*)
- Free (*libre*)

## Overview

- RMarkdown workflow
- Vectors
- Data frames
- Some simple statistics

## Vectors

- Often you want to deal with several numbers at once.
- The `c( )` operator *combines* its contents into a *vector*.

```{r}
c(2,4,6,2,10)
c(2,4,6,2,10) + c(10, 20, 30, 40, 50)
c(2,4,6,2,10) * 1000
```

## Variables

- Often you want to save a vector so you can reuse it.
- The assignment operator `<-` saves a variable.

```{r}
joe <- c(2,4,6,2,10)
joe^2
sum(joe)
mean(joe)
```

## Data Frames in R

- The basic form of data. Think spreadsheet.
- Columns have the same length, but can be different types.
- One row per observational unit.

## Built-in data sets: `iris`

```{r}
data("iris")
summary(iris)
```

## Columns are vectors

Use the `$` operator to return the columns of a data frame.

```{r}
iris$Sepal.Length
```

# *MA-5 in 5 Minutes* {data-background="http://pfeifferpartners.com/wp-content/uploads/2016/03/pfeiffer-westmont-int-winter-main-lobby-stair-M.jpg" data-background-size="contain"}

## Basic Statistical Tests in R

>- Once the data is loaded, most tests are one line of code.
>- Simple graphs are also one-liners (base graphics).

## MA-5 Topics I cover

- One proportion and one mean
- Two proportions and two means (independent samples and paired)
- Multiple proportions ($\chi^2$)
- Multiple means (ANOVA)
- Regression and Correlation
- Multiple linear regression
- MANOVA, Factorial ANOVA, Logistic Regression (optional)

>- *Nonparametric Resampling Techniques for all of the above*

# *One Sample Tests* {data-background="https://cdn.cnn.com/cnnnext/dam/assets/151119194429-04-cancer-dogs-super-169.jpg" data-background-size="contain"}

## One-Proportion Test

$$
H_0: \pi = 0.5 \\ 
H_a: \pi \neq 0.5 \\ 
\hat{p} = 54/83, \quad n=83
$$

```{r, eval=FALSE}
prop.test(54, 83, p = 0.5)
```

## One-Proportion Test

```{r}
prop.test(54, 83, p = 0.5)
```


## One-Sample $t$-test

```{r}
sleep <- read.table("http://math.westmont.edu/ma5/classSleep.txt", header = TRUE)
summary(sleep)
```

## One-Sample $t$-test

$$
H_0: \mu = 0 \\
H_a: \mu \neq 0
$$

```{r, eval=FALSE}
t.test(sleep)
```

## One-Sample $t$-test

```{r}
t.test(sleep)
```

## $t$-test help menu

```{r, eval=FALSE}
?t.test
```

```
Student's t-Test
Description:
     Performs one and two sample t-tests on vectors of data.
Usage:
     t.test(x, ...)
     
     ## Default S3 method:
     t.test(x, y = NULL,
            alternative = c("two.sided", "less", "greater"),
            mu = 0, paired = FALSE, var.equal = FALSE,
            conf.level = 0.95, ...)
```

## One-sided $t$-test

```{r}
t.test(sleep, alternative="less", mu=8)
```

# *Explanatory/Response Scenarios* {data-background="https://pages.wustl.edu/files/pages/olsen/site/slideshow/img3260.jpg" data-background-size="contain"}

## Explanatory/Response variables

| Explanatory | Response | Scenario/Test |
| ---- | ---- | ---- |
| Categorical (2 levels) | Categorical | two-proportion $z$-test |
| Categorical (2 levels, independent) | Quantitative | two-sample $t$-test |
| Categorical (2 levels, paired) | Quantitative | paired $t$-test |
| Categorical (>2 levels) | Categorical | $\chi^2$ test |
| Categorical (>2 levels) | Quantitative | ANOVA $F$-test |
| Quantitative | Quantitative | Regression/Correlation |

## Formulas in R

- Explanatory (x) and response (y) variables are usually denoted using "tilde" notation: `y ~ x`
- Works in different contexts (tests, plots)
- Can also have multiple response, explanatory variables

## Two-sample $t$-test

```{r}
friends <- read.table("http://www.isi-stats.com/isi/data/chap6/CloseFriends.txt", header = TRUE)
summary(friends)
```

## Two-sample $t$-test

```{r, fig.align='center', fig.height=4, fig.width=4}
plot(Friends ~ Sex, data = friends)
```

## Two-sample $t$-test

$$
H_0: \mu_1 - \mu_2 = 0 \\
H_a: \mu_1 - \mu_2 \neq 0 
$$

```{r, eval=FALSE}
t.test(Friends ~ Sex, data = friends)
```

## Two-sample $t$-test

```{r}
t.test(Friends ~ Sex, data = friends)
```

## Paired $t$-test

```{r}
rounding1st <- read.table("http://www.isi-stats.com/isi/data/chap7/FirstBase.txt", header=TRUE)
summary(rounding1st)
```

## Paired $t$-test

$$
H_0: \mu_d  = 0 \\
H_a: \mu_d  \neq 0 
$$

```{r, eval = FALSE}
t.test(rounding1st$narrow, rounding1st$wide, paired = TRUE)
```

## Paired $t$-test

```{r}
t.test(rounding1st$narrow, rounding1st$wide, paired = TRUE)
```

## Chi-square test

```{r}
apunc <- read.table("http://www.isi-stats.com/isi/data/chap8/Acupuncture.txt", header = TRUE)
table(apunc)
head(apunc)
```

## Chi-square test

```{r, fig.align='center', fig.height=4, fig.width=5}
plot(Improvement ~ Acupunture, data = apunc)
```

## Chi-square test

```{r}
chisq.test(table(apunc))
```

## One-way ANOVA

```{r, fig.align='center', fig.height=4, fig.width=5}
diets <- read.table("http://www.isi-stats.com/isi/data/chap9/Diets.txt", header = TRUE)
plot(BMI ~ Diet, data = diets)
```

## One-way ANOVA

```{r}
d.aov <- aov(BMI ~ Diet, data = diets)
anova(d.aov)
```

## Tukey HSD follow-up

```{r}
TukeyHSD(d.aov)
```

## Regression/Corellation

```{r, fig.align='center', fig.height=4, fig.width=5}
textbooks <- read.table("http://www.isi-stats.com/isi/data/chap10/TextbookPrices.txt", header = TRUE)
plot(Price ~ Pages, data = textbooks)
```

## Regression/Corellation

```{r}
summary(lm(Price ~ Pages, data = textbooks))
```

## Correlation Matrix

```{r}
cor(textbooks)
```

## Multiple Regression

## Logistic Regression

## Analysis of Covariance

# *Other scenarios* {data-background="https://imgs.xkcd.com/comics/correlation.png" data-background-size="contain"}

## UCLA Test Chooser

https://stats.idre.ucla.edu/other/mult-pkg/whatstat/

